---
title: "Bayesian Generative Modeling with BGM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Generative Modeling with BGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("BAYESGM_RUN_VIGNETTES"), "true"),
  message = FALSE,
  warning = FALSE
)
```

This vignette shows a complete `BGM` workflow for the
`Sim_heteroskedastic` setting. The code mirrors the Python workflow in
`src/main.py`: generate data with `simulate_z_hetero`, split into train and
test sets, train `BGM`, then impute the held-out response values.

## Basic Running Instructions

The R package is a thin wrapper around the Python `bayesgm` implementation.
Installing the R package from local files, GitHub, or CRAN does **not**
automatically install the Python `bayesgm` package for you.

If you installed the R package from a local checkout of this repository, use
the repository's `src/` directory when installing the Python backend:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python",
  local_path = "/path/to/bayesgm/src"
)

configure_bayesgm(
  python = "/path/to/python",
  pythonpath = "/path/to/bayesgm/src"
)
```

If you installed the R package from GitHub or CRAN and the Python `bayesgm`
package is available from your Python package index, the one-time setup is:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python"
)

configure_bayesgm(
  python = "/path/to/python"
)
```

To run the code while rendering the vignette locally, set:

- `RETICULATE_PYTHON` to the Python interpreter that has the Python `bayesgm` package available
- `BAYESGM_PYTHONPATH` if you want to import the Python package from a source tree
- `BAYESGM_RUN_VIGNETTES=true`

```{r libraries}
library(bayesgm)

configure_bayesgm()
stopifnot(bayesgm_available(configure = FALSE))
```

# Data Generation and Split

```{r data-generation}
sim_data <- load_sim_heteroskedastic(
  n = 20000L,
  z_dim = 10L,
  x_dim = 100L,
  seed = 42L,
  split_seed = 123L
)

data_train <- sim_data$data_train
X_test <- sim_data$X_test
Y_test <- sim_data$Y_test
data_test <- sim_data$data_test

dim(data_train)
dim(data_test)
head(Y_test)
```

# Model Training

```{r model-training}
params <- list(
  dataset = "Sim_heteroskedastic",
  output_dir = tempdir(),
  save_res = TRUE,
  save_model = TRUE,
  use_bnn = FALSE,
  rank = 2L,
  z_dim = 10L,
  x_dim = ncol(data_train),
  lr_theta = 0.005,
  lr_z = 0.005,
  g_units = c(64L, 64L, 64L, 64L, 64L),
  kl_weight = 0.00005,
  lr = 0.001,
  g_d_freq = 1L,
  use_z_rec = TRUE,
  alpha = 0.0,
  gamma = 0.0,
  e_units = c(64L, 64L, 64L, 64L, 64L),
  dz_units = c(64L, 32L, 8L),
  dx_units = c(64L, 32L, 8L)
)

model <- BGM(params = params, random_seed = NULL)

invisible(capture.output({
  model <- model$fit(
    data = data_train,
    epochs = 200L,
    epochs_per_eval = 10L,
    batch_size = 32L,
    use_egm_init = TRUE,
    egm_n_iter = 50000L,
    egm_batches_per_eval = 500L,
    verbose = 1L
  )
}))
```

# Model Prediction

```{r model-prediction}
invisible(capture.output({
  res <- model$predict(
    data = data_test,
    alpha = 0.05,
    bs = 100L,
    n_mcmc = 5000L,
    burn_in = 5000L,
    step_size = 0.01,
    num_leapfrog_steps = 10L,
    seed = 42L
  )
}))

data_imputed <- as.matrix(res$data)
pred_interval <- res$interval
pred_y <- data_imputed[, ncol(data_imputed)]

dim(data_imputed)
dim(pred_interval)
head(pred_y)
```

# Model Evaluation

```{r model-evaluation, fig.width = 7, fig.height = 5}
interval_y <- pred_interval[, 1, ]
lower_y <- interval_y[, 1]
upper_y <- interval_y[, 2]

rmse_y <- sqrt(mean((pred_y - Y_test[, 1])^2))
coverage_y <- mean(Y_test[, 1] >= lower_y & Y_test[, 1] <= upper_y)

data.frame(
  metric = c("RMSE", "Coverage"),
  value = c(rmse_y, coverage_y)
)

plot(
  Y_test[, 1],
  pred_y,
  pch = 16,
  col = grDevices::adjustcolor("steelblue", alpha.f = 0.35),
  xlab = "True held-out response",
  ylab = "Imputed response",
  main = "BGM Imputation on Sim_heteroskedastic"
)
abline(0, 1, col = "black", lwd = 2)
```
