---
title: "Continuous Treatment with CausalBGM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Treatment with CausalBGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("BAYESGM_RUN_VIGNETTES"), "true"),
  message = FALSE,
  warning = FALSE
)
```

This vignette shows a continuous-treatment workflow using the
`Sim_Hirano_Imbens` setting. The code mirrors the Python workflow in
`src/main.py`: load data from the Python sampler, define `params`,
instantiate `CausalBGM(params = params)`, train with `fit()`, then evaluate
the ADRF with `predict()`.

## Basic Running Instructions

The R package is a thin wrapper around the Python `bayesgm` implementation.
Installing the R package from local files, GitHub, or CRAN does **not**
automatically install the Python `bayesgm` package for you.

If you installed the R package from a local checkout of this repository, use
the repository's `src/` directory when installing the Python backend:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python",
  local_path = "/path/to/bayesgm/src"
)

configure_bayesgm(
  python = "/path/to/python",
  pythonpath = "/path/to/bayesgm/src"
)
```

If you installed the R package from GitHub or CRAN and the Python `bayesgm`
package is available from your Python package index, the one-time setup is:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python"
)

configure_bayesgm(
  python = "/path/to/python"
)
```

For a seamless experience in future R sessions, set `RETICULATE_PYTHON` in your
`~/.Renviron`, and set `BAYESGM_PYTHONPATH` as well if you are using a local
repository checkout.

To run the code while rendering the vignette locally, set:

- `RETICULATE_PYTHON` to the Python interpreter that has the Python `bayesgm` package available
- `BAYESGM_PYTHONPATH` if you want to import the Python package from a source tree
- `BAYESGM_RUN_VIGNETTES=true`

```{r libraries}
library(bayesgm)

configure_bayesgm()
stopifnot(bayesgm_available(configure = FALSE))
```

# Data Loading

```{r data-loading}
sim_data <- load_sim_hirano_imbens(N = 1000L, v_dim = 20L, seed = 0L)

x <- sim_data$x
y <- sim_data$y
v <- sim_data$v
grid_val <- seq(0, 3, length.out = 20)
true_effect <- grid_val + 2 * (1 + grid_val)^(-3)

dim(v)
head(x)
head(y)
```

# Model Training

```{r model-training}
params <- list(
  dataset = "Sim_Hirano_Imbens",
  output_dir = tempdir(),
  save_res = FALSE,
  save_model = FALSE,
  binary_treatment = FALSE,
  use_bnn = TRUE,
  z_dims = c(1L, 1L, 1L, 7L),
  v_dim = 20L,
  lr_theta = 0.0001,
  lr_z = 0.0001,
  g_units = c(64L, 64L, 64L, 64L, 64L),
  f_units = c(64L, 32L, 8L),
  h_units = c(64L, 32L, 8L),
  kl_weight = 0.0001,
  lr = 0.0002,
  g_d_freq = 5L,
  use_z_rec = TRUE,
  e_units = c(64L, 64L, 64L, 64L, 64L),
  dz_units = c(64L, 32L, 8L)
)

model <- CausalBGM(params = params, random_seed = NULL)

invisible(capture.output({
  model <- model$fit(
    x = x,
    y = y,
    v = v,
    epochs = 100L,
    epochs_per_eval = 10L,
    batch_size = 32L,
    use_egm_init = TRUE,
    egm_n_iter = 10000L,
    egm_batches_per_eval = 500L,
    verbose = 1L
  )
}))
```

# Model Prediction

```{r model-prediction}
invisible(capture.output({
  res <- model$predict(
    x = x,
    y = y,
    v = v,
    alpha = 0.01,
    x_values = grid_val,
    n_mcmc = 3000L,
    burn_in = 3000L,
    q_sd = 1.0
  )
}))

pred_effect <- as.numeric(res$effect)
pred_interval <- as.matrix(res$interval)
pred_lower <- pred_interval[, 1]
pred_upper <- pred_interval[, 2]

length(pred_effect)
head(pred_effect)
head(pred_interval)
```

# Model Evaluation

```{r model-evaluation, fig.width = 7, fig.height = 5}
rmse <- sqrt(mean((true_effect - pred_effect)^2))
mape <- mean(abs((true_effect - pred_effect) / true_effect))

data.frame(
  metric = c("RMSE", "MAPE"),
  value = c(rmse, mape)
)

cat(sprintf(
  "RMSE (Root Mean Squared Error): %.4f\n",
  rmse
))
cat(sprintf(
  "MAPE (Mean Absolute Percentage Error): %.4f\n",
  mape
))

plot(
  grid_val,
  true_effect,
  type = "n",
  ylim = range(c(true_effect, pred_lower, pred_upper), finite = TRUE),
  xlab = "Treatment level",
  ylab = "ADRF",
  main = "Predicted and True ADRF"
)
polygon(
  x = c(grid_val, rev(grid_val)),
  y = c(pred_lower, rev(pred_upper)),
  border = NA,
  col = grDevices::adjustcolor("firebrick", alpha.f = 0.2)
)
lines(grid_val, true_effect, col = "black", lwd = 2)
lines(grid_val, pred_effect, col = "firebrick", lwd = 2)
legend(
  "bottomright",
  legend = c("True ADRF", "Predicted ADRF", "Predicted 95% interval"),
  col = c("black", "firebrick", NA),
  lty = c(1, 1, NA),
  lwd = c(2, 2, NA),
  pch = c(NA, NA, 15),
  pt.cex = c(NA, NA, 2),
  pt.bg = c(NA, NA, grDevices::adjustcolor("firebrick", alpha.f = 0.2)),
  bty = "n"
)
```
