---
title: "Binary Treatment with CausalBGM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary Treatment with CausalBGM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("BAYESGM_RUN_VIGNETTES"), "true"),
  message = FALSE,
  warning = FALSE
)
```

This vignette shows a complete binary-treatment workflow with `CausalBGM`.
The code mirrors the Python workflow in `src/main.py` for the `Semi_acic`
setting: define `params`, instantiate `CausalBGM(params = params)`, train with
`fit()`, and estimate effects with `predict()`.

## Basic Running Instructions

The R package is a thin wrapper around the Python `bayesgm` implementation.
Installing the R package from local files, GitHub, or CRAN does **not**
automatically install the Python `bayesgm` package for you.

If you installed the R package from a local checkout of this repository, use
the repository's `src/` directory when installing the Python backend:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python",
  local_path = "/path/to/bayesgm/src"
)

configure_bayesgm(
  python = "/path/to/python",
  pythonpath = "/path/to/bayesgm/src"
)
```

If you installed the R package from GitHub or CRAN and the Python `bayesgm`
package is available from your Python package index, the one-time setup is:

```r
library(bayesgm)

install_bayesgm_python(
  python = "/path/to/python"
)

configure_bayesgm(
  python = "/path/to/python"
)
```

For a seamless experience in future R sessions, set `RETICULATE_PYTHON` in your
`~/.Renviron`, and set `BAYESGM_PYTHONPATH` as well if you are using a local
repository checkout.

This vignette uses the bundled dataset `acic_2018_example`, so no separate data
download is required once the R package is installed.

To run the code while rendering the vignette locally, set:

- `RETICULATE_PYTHON` to the Python interpreter that has the Python `bayesgm` package available
- `BAYESGM_PYTHONPATH` if you want to import the Python package from a source tree
- `BAYESGM_RUN_VIGNETTES=true`

```{r libraries}
library(bayesgm)

configure_bayesgm()
stopifnot(bayesgm_available(configure = FALSE))
```

# Data Loading

```{r data-loading}
data("acic_2018_example", package = "bayesgm")

ufid <- acic_2018_example$ufid
x <- acic_2018_example$x
y <- acic_2018_example$y
v <- acic_2018_example$v
ite_true <- acic_2018_example$ite_true
ate_true <- acic_2018_example$ate_true

ufid
dim(v)
head(x)
head(y)
```

# Model Training

```{r model-training}
params <- list(
  dataset = "Semi_acic",
  output_dir = tempdir(),
  save_res = TRUE,
  save_model = TRUE,
  binary_treatment = TRUE,
  use_bnn = TRUE,
  z_dims = c(3L, 6L, 3L, 6L),
  v_dim = 177L,
  lr_theta = 0.0001,
  lr_z = 0.0001,
  g_units = c(64L, 64L, 64L, 64L, 64L),
  f_units = c(64L, 32L, 8L),
  h_units = c(64L, 32L, 8L),
  kl_weight = 0.0001,
  lr = 0.0002,
  g_d_freq = 5L,
  use_z_rec = TRUE,
  e_units = c(64L, 64L, 64L, 64L, 64L),
  dz_units = c(64L, 32L, 8L)
)

model <- CausalBGM(params = params, random_seed = NULL)

invisible(capture.output({
  model <- model$fit(
    x = x,
    y = y,
    v = v,
    epochs = 100L,
    epochs_per_eval = 10L,
    batch_size = 32L,
    use_egm_init = TRUE,
    egm_n_iter = 10000L,
    egm_batches_per_eval = 500L,
    verbose = 1L
  )
}))
```

# Model Prediction

```{r model-prediction}
invisible(capture.output({
  res <- model$predict(
    x = x,
    y = y,
    v = v,
    alpha = 0.01,
    n_mcmc = 1000L,
    burn_in = 1000L,
    q_sd = 1.0
  )
}))

pred_ite <- as.numeric(res$effect)
pred_interval <- as.matrix(res$interval)

length(pred_ite)
head(pred_ite)
head(pred_interval)
```

# Model Evaluation

For this ACIC example, the package dataset includes the ground-truth
counterfactual outcomes, so the true individual treatment effects and average
treatment effect are available directly.

```{r model-evaluation, fig.width = 7, fig.height = 5}
pred_ate <- mean(pred_ite)
delta_ate <- abs(pred_ate - ate_true)
delta_pehe <- mean((pred_ite - ite_true)^2)

data.frame(
  metric = c("True ATE", "Predicted ATE", "Delta ATE", "Delta PEHE"),
  value = c(ate_true, pred_ate, delta_ate, delta_pehe)
)

cat(sprintf(
  "Delta ATE (Absolute Error in Average Treatment Effect): %.4f\n",
  delta_ate
))
cat(sprintf(
  "Delta PEHE (Precision in Estimation of Heterogeneous Effect): %.4f\n",
  delta_pehe
))

plot(
  ite_true,
  pred_ite,
  pch = 16,
  col = grDevices::adjustcolor("firebrick", alpha.f = 0.4),
  xlab = "True ITE",
  ylab = "Predicted ITE",
  main = "ITE Evaluation"
)
abline(0, 1, col = "black", lwd = 2)

```
